Composable data visualizations

### 2020-05-23 wip-compost/paper.tex
Implementation, working on references

\NeedsTeXFormat{LaTeX2e}
\documentclass{jfp}
\usepackage{xcolor}

@@

\bibliographystyle{jfp}

@@

\begin{figure}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{rcl}
l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
\end{array}
\end{equation*}
\vspace{-1em}
\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
\label{fig:scale}
\vspace{-1em}
\end{figure}
\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
  & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
\label{fig:control}
\end{figure}

@@

\end{figure}
\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{mouseUp}~(\lambda \,x~y \rightarrow e), s  &|& \kvd{mouseDown}~(\lambda \,x~y\rightarrow e), s  \\
    &|& \kvd{mouseMove}~(\lambda \,x~y \rightarrow e), s &|& (\ldots)
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Additional combinators for mouse-based interaction, extending earlier definition of $s$.}
\label{fig:mouse}
\end{figure}

@@

The next part of the visualization is the \ident{update} function which takes an old state together
with an event and produces a new state:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =
  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
\hspace{1.05em}~\ident{update}~(m,s)~(\ident{Update}(\_, \_)) = m,s\\
\hspace{1.05em}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
The first two cases handle the \ident{Update} event. The event carries two values, $p, v$, which
represent the party (which bar the user is dragging) and the new value (new number of seats).
If the user is currently dragging, we replace the value associated with the party $p$ in the
list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ingored.
The last case handles the \ident{Moving} event, which simply replaces the first component of the
state tuple, i.e.~a flag indicating whether a mouse button is down.

\begin{figure}[t]
  \includegraphics[scale=0.57]{figures/youdraw}
  \vspace{0.25em}
  \caption{Interactive ``You draw it'' data visualization. The user moves cursor to a bar (left),
    pushes a mouse button and drags the bar to the position that they think is the correct one (right).}
  \label{fig:youdraw}
\end{figure}

@@

\section{Implementation structure}
\label{sec:impl}

The Compost library is open-source and the full source code, implmented in the functional
language F\#, can be found at \urrl{http://github.com/compostjs}.
As is often the case with functional domain-specific languages, the implementation is easy once
we find the right collection of basic primitives. This largely applies to the Compost library
and so we will not go into the implementation details. However, it is worth giving an
outline of the implmentation structure.

@@

\begin{equation*}
\begin{array}{rclcl}
s &=& \kvd{scaledNest}_x~v_{min},\;v_{max},\;s_x,\;s &|& (\ldots)
\end{array}
\end{equation*}

\begin{equation*}
\ident{calculateScales} ~:~ \ident{Shape} \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}
\end{equation*}

\begin{equation*}
\begin{array}{l}
\ident{calculateScales}~(\kvd{nest}_x~v_{min},\;v_{max},\;s) = \\
\quad\fkvd{let}~(s_x, s_y),s' = \ident{calculateScales}~s\\
\quad(\ident{union}~(\ident{singleton}~v_{min})~(\ident{singleton}~v_{max}), s_y),~\kvd{scaledNest}_x~v_{min},\;v_{max},\;s_x,\;s'\\
\\
\ident{calculateScales}~(\kvd{overlay}~l) = \\
\quad\fkvd{let}~\ident{scales},\;l'~=~\ident{unzip}~(\ident{map}~\ident{calculateScales}~l)\\
\quad\fkvd{let}~s_x,s_y~=~\ident{unzip}~\ident{scales}\\
\quad(\ident{reduce}~\ident{union}~s_x,~\ident{reduce}~\ident{union}~s_y), \kvd{overlay}~l'
\end{array}
\end{equation*}

\subsection{Drawing shapes}

\begin{equation*}
\ident{project}~:~\ident{float}*\ident{float}\rightarrow \ident{Scale}\rightarrow\ident{Value} \rightarrow \ident{float}
\end{equation*}

\begin{equation*}
\ident{drawShape}~:~\ident{Area} \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}\rightarrow\ident{Svg}
\end{equation*}

\begin{equation*}
\begin{array}{l}
\ident{drawShape}~(x_1, y_1, x_2, y_2)~(s_x, s_y)~(\kvd{scaledNest}_x~v_{min},\;v_{max},\;{is}_x,\;\ident{shape}) = \\
\quad\fkvd{let}~x_1' = \ident{project}~(x_1, x_2)~s_x~v_{min}\\
\quad\fkvd{let}~x_2' = \ident{project}~(x_1, x_2)~s_x~v_{max}\\
\quad\ident{drawShape}~(x_1', y_1, x_2', y_2)~({is}_x, s_y)~\ident{shape}\\
\\
\ident{drawShape}~a~s~(\kvd{overlay}~l) = \\
\quad\ident{append}~(\ident{map}~(\ident{drawShape}~a~s)~l)

\end{array}
\end{equation*}

a

b

c

d

e

f

g

h

i
\newpage

\subsection*{Controlling and nesting scales}

\subsection*{Nesting of charts and scales}

~
~
\noindent
[TODO: We really need illustrations for those example charts, but that would be more work!]
~
[TODO: Maybe use this chart as a motivation: https://wellcomeopenresearch.org/articles/4-63/v1]
~
\section{Random ideas}
\begin{itemize}
\item We could define a type system to catch categorical/continuous value mismatch in a single %scale.
\item It would be worth thinking about ordinal values, which are categorical but can be sorted.
\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
 or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
 every value (including bar chart colors) should be coming from some scale...
\item Implementing something like \kvd{axis} or \kvd{roundScale} as an actual derived primitive is
 a bit tricky, because it needs to invoke a part of the normal rendering workflow (to run the
 automatic inference of scales on the nested shape) -- this might be just implementation issue,
 but it could be some more basic problem.
\end{itemize}
\end{document}

%

% \usepackage{booktabs}
% \usepackage{subcaption}
% \usepackage{lineno,hyperref,xcolor}
% \usepackage{flushend}
% \usepackage{stmaryrd}
% \usepackage{amssymb}
% \usepackage{xypic}
% \usepackage{semantic}
% \usepackage{booktabs}
% \usepackage{enumitem}
% \usepackage{enumerate}
% \usepackage{amsmath}
% \usepackage[T1]{fontenc}
%
% \setlist{leftmargin=6mm}
% \newcounter{thc}
% \newcounter{dfc}
%
% \theoremstyle{plain}
% \newtheorem{lem}[thc]{Lemma}
% \newtheorem{theorem}[thc]{Theorem}
%
% \theoremstyle{definition}
% \newtheorem{axiom}[dfc]{Axiom}
% \newtheorem{definition}[dfc]{Definition}
%
% \title{Compost: Library for composable data visualization}
% %\author{Tomas Petricek}
% %\affiliation{
% %  \institution{University of Kent}
% %  \country{United Kingdom}
% %}
% %\email{tomas@tomasp.net}
%
% %\begin{CCSXML}
% %<ccs2012>
% %<concept>
% %<concept_id>10011007.10011006.10011008</concept_id>
% %<concept_desc>Software and its engineering~General programming languages</concept_desc>
% %<concept_significance>500</concept_significance>
% %</concept>
% %<concept>
% %<concept_id>10003456.10003457.10003521.10003525</concept_id>
% %<concept_desc>Social and professional topics~History of programming languages</concept_desc>
% %<concept_significance>300</concept_significance>
% %</concept>
% %</ccs2012>
% %\end{CCSXML}
% %
% %\ccsdesc[500]{Software and its engineering~General programming languages}
% %\ccsdesc[300]{Social and professional topics~History of programming languages}
%
% \definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
% \definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
% \definecolor{idclr}{rgb}{0.0,0.0,0.0}
% \definecolor{numclr}{rgb}{0.0,0.4,0.0}
% \definecolor{strclr}{rgb}{0.4,0.4,0.0}
% \definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
% \definecolor{prepclr}{rgb}{0.6,0.0,0.2}
% \newcommand{\vect}[1]{\langl #1 \rangl}
% \newcommand{\langl}{\begin{picture}(4.5,7)
% \put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
% \put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
% \end{picture}}
% \newcommand{\rangl}{\begin{picture}(4.5,7)
% \put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
% \put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
% \end{picture}}
% \newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
% \newcommand{\lsep}{\;\;|\;\;}
% \newcommand{\num}[1]{\textcolor{numclr}{#1}}
% \newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
% \newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
% \newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
% \newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
% \newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
% \newcommand{\dom}{\ident{dom}}
% \newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
%
% \newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
% \newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
% \newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
% \newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
% \newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
% \newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
% \newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
% \newcommand{\rname}[1]{{\sffamily\small(#1)}}
% \newcommand{\ername}[1]{\vspace{0.75em}\rname{#1}\hspace{0.5em}}
% \newcommand{\preview}[1]{\,\textnormal{\guillemotleft} #1\textnormal{\guillemotright}\,}
%
% \begin{document}
% %\begin{abstract}
% %\end{abstract}
% \maketitle
%
% % ==================================================================================================
%

>>>>

\NeedsTeXFormat{LaTeX2e}
\documentclass{jfp}
\usepackage{amssymb}
\usepackage{xcolor}

@@

@@

\begin{figure}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{rcl}
l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
\end{array}
\end{equation*}
\vspace{-1em}
\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
\label{fig:scale}
\vspace{-1em}
\end{figure}
\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
  & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
\end{array}
\end{equation*}
\caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
\label{fig:control}
\end{figure}

@@

\end{figure}
\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{mouseUp}~(\lambda \,x~y \rightarrow e), s  &|& \kvd{mouseDown}~(\lambda \,x~y\rightarrow e), s  \\
    &|& \kvd{mouseMove}~(\lambda \,x~y \rightarrow e), s &|& (\ldots)
\end{array}
\end{equation*}
\caption{Additional combinators for mouse-based interaction, extending earlier definition of $s$.}
\vspace{0.5em}
\label{fig:mouse}
\end{figure}

@@

The next part of the visualization is the \ident{update} function which takes an old state together
with an event and produces a new state:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{update}~(\_,s)~(\ident{Moving}(m)) = m,s\\
\hspace{1.05em}~\ident{update}~(\fkvd{true},s)~(\ident{Update}(p, v)) =\\
\hspace{1.05em}\qquad  \fkvd{true}, \ident{map}~(\lambda(k, o) \rightarrow k, \fkvd{if}~k = p~\fkvd{then}~v~\fkvd{else}~o)~s\\
\hspace{1.05em}~\ident{update}~(m,s)~(\ident{Update}(\_, \_)) = m,s\\
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
The first two cases handle the \ident{Update} event. The event carries two values, $p, v$, which
represent the party (which bar the user is dragging) and the new value (new number of seats).
If the user is currently dragging, we replace the value associated with the party $p$ in \newpage \noindent the
list $s$ using the \ident{map} function. If the user is not currently dragging, the event is ingored.
The last case handles the \ident{Moving} event, which simply replaces the first component of the
state tuple, i.e.~a flag indicating whether a mouse button is down.

\begin{figure}[t]
  \vspace{0.5em}
  \includegraphics[scale=0.57]{figures/youdraw}
  \vspace{1em}
  \caption{Interactive ``You draw it'' data visualization. The user moves cursor to a bar (left),
    pushes a mouse button and drags the bar to the position that they think is the correct one (right).}
  \label{fig:youdraw}
\end{figure}

@@

\section{Implementation structure: Scale inference and projection}
\label{sec:impl}

Compost is an open-source library, implmented in the functional language F\#. The full source
code can be found at \urrl{http://github.com/compostjs}.
As is often the case with functional domain-specific languages, the implementation is easy once
we find the right collection of basic primitives and the right structure for the implementation.
This largely applies to the Compost library and so we will not go into the implementation details.
It is, however, worth giving an outline of the implmentation structure.

@@

In order to render a shape, we need to know the range of values that should appear on the X and
Y axes. This is done by inferring a scale for each of the axes from the individual X and Y
coordinates that specify shape locations. As discussed earlier, a scale can be either categorical
(displaying only categorical values) or contiuous (displaying only continuous values). When
inerring scales, we use two helper operations; \ident{union}, discussed earlier, combines two
scales and \ident{singleton} creates a scale from a single coordinate.

The operation that infers the scales of a shape is \ident{calculateScales}. It takes a shape
and produces a pair of X and Y scales, together with a transformed shape:
%
\begin{equation*}
  \ident{calculateScales} ~:~ \ident{Shape} \rightarrow (\ident{Scale}*\ident{Scale}) * \ident{Shape}
\end{equation*}

\vspace{-1em}
\noindent
The operation does not need to transform the shape in most cases. The exception is the
shape $\kvd{nest}_{x/y}~v_{min}, v_{max}, s$. In this case, the returned scale is based solely on the
values of $v_{min}$ and $v_{max}$. For rendering we need to keep the inferred scales of the nested
shape $s$. To do so, the operation replaces the $\kvd{nest}_{x/y}$ shape with an auxiliary shape
$\kvd{scaledNest}_{x/y}$:
%
\begin{equation*}
\begin{array}{rclcl}
s &=& \kvd{scaledNest}_{x/y}~v_{min},\;v_{max},\;s_{x/y},\;s &|& (\ldots)
\end{array}
\end{equation*}

\vspace{-1em}
\noindent
There are two kinds of cases handled by \ident{calculateScales}. For primitives, it constructs
a pair of scales from individual coordinates using \ident{union} and \ident{singleton}. For
shapes containing a sub-shape, the operation calculates the scales of a sub-shape recursively
and then adapts those somehow. To illustrate, we consider two interesting cases:
%
\begin{equation*}
\begin{array}{l}
\ident{calculateScales}~(\kvd{nest}_x~v_{min},\;v_{max},\;s) = \\
\quad\fkvd{let}~(s_x, s_y),s' = \ident{calculateScales}~s\\
\quad(\ident{union}~(\ident{singleton}~v_{min})~(\ident{singleton}~v_{max}), s_y),~\kvd{scaledNest}_x~v_{min},\;v_{max},\;s_x,\;s'\\
\\
\ident{calculateScales}~(\kvd{overlay}~l) = \\
\quad\fkvd{let}~\ident{scales},\;l'~=~\ident{unzip}~(\ident{map}~\ident{calculateScales}~l)\\
\quad\fkvd{let}~s_x,s_y~=~\ident{unzip}~\ident{scales}\\
\quad(\ident{reduce}~\ident{union}~s_x,~\ident{reduce}~\ident{union}~s_y), \kvd{overlay}~l'
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
When calculating the scales of the $\kvd{nest}_x$, the function first calculate scales of the
sub-shape $s$ recursively. The resulting Y scale $s_y$ is returned as the result, while the X
scale is obtained from the two coordinates $v_{min}$ and $v_{max}$. This is also the case where
the shape is transformed and the returned $\kvd{scaledNest}_x$ shape stores the inferred X scale
$s_x$ of the sub-shape $s$. The second example is the $\kvd{overlay}$ case which recursively
infers scales of all sub-shapes and combines those using the list folding function
\ident{reduce} with $\ident{union}$ as an argument.

\subsection{Projecting coordinates and drawing}
The key operation that needs to be performed when drawing a shape is projecting coordinates from
domain-specific values to the screen coordinates. As we draw a shape, we keep the X and Y scale
and the space in pixels that it should be drawn on. Initially, the X and Y scales are those
inferred for the entire shape and the space in pixels is $\num{0}\ldots\ident{width}$ and
$\num{0}\ldots\ident{height}$ where $\ident{width}\times\ident{height}$ is the size of the target
SVG element.

The key calculation is done by the \ident{project} function, which takes the space in pixels (as a
pair of floating-point numbers representing the range), the current scale and a domain-specific value
and produces a coordinate in pixels:
%
\begin{equation*}
\ident{project}~:~\ident{float}*\ident{float}\rightarrow \ident{Scale}\rightarrow\ident{Value} \rightarrow \ident{float}
\end{equation*}

\vspace{-1.25em}
\noindent
The function is only defined if the value and scale are compatible. If both are continuous, the
function performs a simple linear transformation. If both are categorical, the available pixel
space is divided into a equally-sized bins, one for each categorical value on the scale, and the
value is then projected into the appropriate bin.

The drawing of shapes is done by a function that takes the available area as a quadruple $(x_1, y_1), (x_2, y_2)$
together with the X and Y scale corresponding to the area and a shape to be drawn. The result is
a data structure representing a SVG document:
%
\begin{equation*}
\ident{drawShape}~:~(\ident{float}*\ident{float})*(\ident{float}*\ident{float}) \rightarrow \ident{Scale}*\ident{Scale}\rightarrow\ident{Shape}\rightarrow\ident{Svg}
\end{equation*}

\vspace{-1.25em}
\noindent
For primitive shapes, the operation projects the coordinates using \ident{project} and construct
a corresponding SVG document. For shapes with sub-shapes it calls itself recursively, possibly
with an adjusted scale or area. The two cases discussed earlier illustrate this:
%
\begin{equation*}
\begin{array}{l}
\ident{drawShape}~a~s~(\kvd{overlay}~l) = \\
\quad\ident{concat}~(\ident{map}~(\ident{drawShape}~a~s)~l)\\
\\[-0.5em]
\ident{drawShape}~((x_1, y_1), (x_2, y_2))~(s_x, s_y)~(\kvd{scaledNest}_x~v_{min},\;v_{max},\;{ns}_x,\;\ident{shape}) = \\
\quad\fkvd{let}~x_1' = \ident{project}~(x_1, x_2)~s_x~v_{min}\\
\quad\fkvd{let}~x_2' = \ident{project}~(x_1, x_2)~s_x~v_{max}\\
\quad\ident{drawShape}~((x_1', y_1), (x_2', y_2))~({ns}_x, s_y)~\ident{shape}\\
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
When drawing \kvd{overlay}, the function draws all sub-shapes onto the same
area using the same scales and then concatenates the returned SVG components the \ident{concat} helper.
The $\kvd{scaledNest}_x$ case is more illuminating. Here, we first
use \ident{project} to find the range $x_1', x_2'$ corresponding to the domain-values
$v_{min},v_{max}$. This defines the area corresponding to the nested scale ${ns}_x$, onto
which the X coordinates in the sub-shape $\ident{shape}$ should be projected.
To do this, we recursively call \ident{drawShape}, but use $x_1'$ and $x_2'$ as the X coordinates
of the target area and ${ns}_x$ as the X scale. The Y area and scales are propagated unchanged.

\newpage

\section{Conclusions}
\begin{itemize}
\item We could define a type system to catch categorical/continuous value mismatch in a single %scale.
\item It would be worth thinking about ordinal values, which are categorical but can be sorted.
\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
 or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
 every value (including bar chart colors) should be coming from some scale...
\item Implementing something like \kvd{axis} or \kvd{roundScale} as an actual derived primitive is
 a bit tricky, because it needs to invoke a part of the normal rendering workflow (to run the
 automatic inference of scales on the nested shape) -- this might be just implementation issue,
 but it could be some more basic problem.
\end{itemize}

\cite{gcharts}
\cite{seaborn}
\cite{youdraw}
\cite{grammar}
\cite{layered}
\cite{vega-lite}
\cite{vega}
\cite{d3}
\cite{polaris}

\bibliographystyle{jfp}
\bibliography{paper}

\end{document}
%
